---
title: "Getting My Colleagues Hooked on R"
author: "Florian Privé"
date: "August 5, 2016"
output:
  ioslides_presentation:
    css: styles.css
    self_contained: yes
---

```{r, include=FALSE}
pacman::p_load(googlesheets, gsubfn, stringr)
printf <- function(...) cat(sprintf(...))

sheet <- gs_key("113h74NVW_VZWuF9EttiYTNVONS-SXEvi7MwGizZD2LE")

sheet.content <- gs_read(sheet)

responses <- sheet.content[, 2]
n <- nrow(responses)
```


## What `r n` of you wanted to learn

```{r, echo=FALSE, results='asis'}
url <- "https://docs.google.com/forms/d/13Hrr9x-3ASaPKPJGRpKzsGWlkwu59LGITtw1ldG7FAU/edit#responses"
url <- url(url)
url.lines <- readLines(url, encoding = "UTF-8", warn = FALSE)
pattern <-  "\\[\"([^\"]*)\",,,,0\\]"
questions <- unlist(gsubfn::strapply(url.lines, pattern))

counts <- str_count(responses, coll(questions))

counts.lvl <- setdiff(sort(unique(counts), decreasing = TRUE), 0)

for (n in counts.lvl) {
  if (n == 2) printf("\n***\n")
  printf("- for **%d** of you:\n", n)
  q.tmp <- questions[counts == n]
  for (q in q.tmp) {
    printf("    - %s\n", q)
  }
}
```

<!-- ## What is R?  -->

<!-- <iframe width="640" height="360" src="https://www.youtube.com/embed/TR2bHSJ_eck" frameborder="0" allowfullscreen></iframe> -->


## Some facts about the growth of R:

- R is #5 of all programming languages ([IEEE Spectrum, July 2016](https://www.r-bloggers.com/r-moves-up-to-5th-place-in-ieee-language-rankings/))

```{r, fig.align='center', echo=FALSE}
knitr::include_graphics("http://revolution-computing.typepad.com/.a/6a010534b1db25970b01bb092485d1970d-800wi")
```

<!-- The other four languages in the top 5 (C, Java, Python amd C++) are all general-purpose languages, suitable for just about any programming task. R by contrast is a language specifically for data science, and its high ranking here reflects both the critical importance of data science as a discipline today, and of R as the language of choice for data scientists. -->

---

```{r, echo = FALSE}
thepage <- readLines('https://cran.r-project.org/web/packages/')
theline <- grep("Currently, the CRAN package repository features [0-9]+ available packages.", thepage)
thesplit <- strsplit(thepage[theline], split = " ")[[1]]
thepos <- match("available", thesplit) - 1
n <- thesplit[thepos]
```

- There are now  `r n` available packages on CRAN ([CRAN: Contributed Packages, `r Sys.Date()`](https://cran.r-project.org/web/packages/))

```{r, fig.align='center', echo=FALSE, out.height=500}
knitr::include_graphics("http://a3.typepad.com/6a017d41eeee1a970c01bb08ef2103970d-pi")
```


## Manipulating data? Ask Hadley Wickham!

R packages that he has developped (from [his website](http://hadley.nz/)):

- Data science
    - <a href="http://ggplot2.org">ggplot2</a> for visualising data.
    - <a href="http://github.com/hadley/dplyr">dplyr</a> for manipulating data.
    - <a href="http://github.com/hadley/tidyr">tidyr</a> for tidying data.
    - <a href="http://github.com/hadley/stringr">stringr</a> for working with strings.
    - <a href="http://github.com/hadley/lubridate">lubridate</a> for working with date/times.
    
---

- Data import
    - <a href="http://github.com/hadley/readr">readr</a> for reading .csv and fwf files.
    - <a href="http://github.com/hadley/readxl">readxl</a> for reading .xls and .xlsx files.
    - <a href="http://github.com/hadley/haven">haven</a> for SAS, SPSS, and Stata files.
    - <a href="http://github.com/hadley/httr">httr</a> for talking to web APIs.
    - <a href="http://github.com/hadley/rvest">rvest</a> for scraping websites.
    - <a href="http://github.com/hadley/xml2">xml2</a> for importing XML files.
    
- Software engineering
    - <a href="http://github.com/hadley/devtools">devtools</a> for general package development.
    - <a href="http://github.com/klutometis/roxygen">roxygen2</a> for in-line documentation.
    - <a href="http://github.com/hadley/testthat">testthat</a> for unit testing.

## R: more than just statistics

R was orginally a language __written by statisticians for statisticians__.

Now, it adds more and more features.

## Elegant visualizations tools
- [ggplot2](http://ggplot2.org/)

```{r, echo=TRUE, out.height=330, out.width=530}
library(ggplot2)
dsamp <- diamonds[sample(nrow(diamonds), size = 1000), ]
p <- qplot(carat, price, data = dsamp, colour = clarity)
print(p)
```

## Interactive visualizations tools
- [plotly](https://plot.ly/ggplot2/): interactive plots with the syntax of ggplot2

```{r, eval=FALSE}
library(plotly)
ggplotly(p)
```
<center>
```{r, message=FALSE, out.height=380, out.width=580, echo=FALSE}
library(plotly)
ggplotly(p)
```
</center>

## Use of faster C++ code when needed

> - easy use or C++ code with [Rcpp](https://cran.r-project.org/web/packages/Rcpp/), or other languages
> - advanced (and fast) manipulation of datasets:
    - [data.table](https://cran.r-project.org/web/packages/data.table/): for instance, it enables in-place changes.
    - [dplyr](https://cran.r-project.org/web/packages/dplyr/): fast manipulations of data.frame-like objects.
> - easy parallel computing: [foreach](https://cran.r-project.org/web/packages/foreach/)
> - easy use of on-disk matrices: [bigmemory](https://cran.r-project.org/web/packages/bigmemory/)

You may use JavaScript or C++ code without knowing it.

## Example with foreach and bigmemory

> - Say you have:
    - A SNP big.matrix X stored on-disk in directory _backingfiles_,
    - Infos on the positions of the SNPs (the first 40,000 SNPs are in chromosome 1, then 38,000 are in chromosome 2, etc.),

> - And you have to do some computations which are independent with respect to chromosomes. You want to use __Parallel Computing__!

> - How to do use Parallel Computing on massive genotype matrices?

***

```{r, eval = FALSE, out.height=300}
DO_all <- function(X, infos, ncores) {
  DO_chr <- function(X.desc, lims) {
    X.chr <- sub.big.matrix(X.desc, 
                          firstCol = lims[1], 
                          lastCol = lims[2],
                          backingpath = "backingfiles")
    ## Do something with X.chr
  }
  range.chr <- LimsChr(infos)
  X.desc <- describe(X)
  obj <- foreach(chr = 1:nrow(range.chr), 
                 .packages = "bigmemory")
  expr_fun <- function(chr) {
    DO_chr(X.desc, range.chr[chr, ])
  }
  res <- foreach2(obj, expr_fun, ncores)
}
```

***
```{r, eval = FALSE}
LimsChr <- function(infos) {
  map.rle <- rle(infos$map$chromosome)
  upper <- cumsum(map.rle$length)
  lower <- c(1, upper[-length(upper)] + 1)
  cbind(lower, upper, "chr" = map.rle$values)
}

foreach2 <- function(obj, expr_fun, ncores) {
  if (is.seq <- (ncores == 1)) {
    foreach::registerDoSEQ()
  } else {
    cl <- parallel::makeCluster(ncores)
    doParallel::registerDoParallel(cl)
  }
  res <- eval(parse(
    text = sprintf("foreach::`%%dopar%%`(obj, expr_fun(%s))",
                   obj$argnames)))
  if (!is.seq) parallel::stopCluster(cl)
  return(res)
}
```

## We have RStudio

Live demo

## References and further reading

- [7 Tips For Getting Your Colleagues Hooked on R](http://scl.io/QZxZZl6u#gs.zMhz76Q)
- [Video: What is R?](https://www.youtube.com/embed/TR2bHSJ_eck)
- [How Companies Use R to Compete in a Data-Driven World](http://data-informed.com/companies-use-r-compete-data-driven-world/)
- [How the growth of R helps data-driven organizations succeed](http://www.slideshare.net/RevolutionAnalytics/how-the-growth-of-r-helps-datadriven-organizations-succeed)
- [A segmented model of CRAN package growth](https://www.r-bloggers.com/a-segmented-model-of-cran-package-growth/)
