---
title: "What R can do - Grenoble RUG"
output:
  xaringan::moon_reader:
    seal: false
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', dev = "svg", comment = "")
options(width = 70)
```

class: title-slide center middle inverse

# What `r icon::fa("r-project")` can do for you

<br>

## Florian Privé

### Grenoble RUG - September 13, 2018

---

## Contents

- Statistics

- Visualization

- Data Science

- High Performance Computing

- Web

- Reporting

- RStudio IDE

- Community

- Learn R

---

# Data Science

---

## Data manipulation with {dplyr}

```{r, message=FALSE}
library(dplyr)
(flights <- nycflights13::flights)
```

---

R package {dplyr} aims to provide a function for each basic verb of data manipulation:

- ``filter()`` (and ``slice()``)

- `arrange()`

- `select()` (and `rename()`)

- `distinct()`

- `mutate()` (and `transmute()`)

- `summarise()`

- and many others..

---

---

## Filter

```{r}
filter(flights, month == 1, day == 1)
```

---

```{r}
arrange(flights, desc(dep_delay))
```

---

```{r}
mutate(flights, gain = arr_delay - dep_delay,
  speed = distance / air_time * 60)
```

---

```{r}
flights2 <- flights %>%
  filter(month == 1, day == 1) %>%
  arrange(desc(dep_delay)) %>%
  mutate(gain = arr_delay - dep_delay,
         speed = distance / air_time * 60)
print(flights2, n = 6)
```

---

## Where to learn R?



---

Can't know everything -> need community (grrr [Join the French-speaking R community](https://join.slack.com/t/r-grrr/shared_invite/enQtMzI4MzgwNTc4OTAxLWZlOGZiZTBiMWU0NDQ3OTYzOGE1YThiODgwZWNhNWEyYjI4ZDJiNmNhY2YyYWI5YzFiOTFkNDYxYzkwODUwNWM))
Learn more with [this blog post](https://masalmon.eu/2018/07/22/wheretogethelp/)

https://cran.r-project.org/web/views/
https://github.com/rstudio/RStartHere

- data analysis + stats + DL + DB

- web scraping, shiny apps

- HPC

- RStudio 

- visualization (ggplot, gganimate, rayshader)

- GIS

- Time series / Text / Images

- learning resources

- R Markdown

---


---

class: inverse, center, middle

# Thanks!

<br/><br/>

<br/>

`r icon::fa("twitter")` [privefl](https://twitter.com/privefl) &nbsp;&nbsp;&nbsp;&nbsp; `r icon::fa("github")` [privefl](https://github.com/privefl) &nbsp;&nbsp;&nbsp;&nbsp; `r icon::fa("stack-overflow")` [F. Privé](https://stackoverflow.com/users/6103040/f-priv%c3%a9)

.footnote[Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan).]
